---
title: Basic BaSTA
layout: default
output: bookdown::html_chapter
---
<!---
1. Use standard R Markdown.
2. Headings with "##" will be automatically turned into table of contents. Therefore, for longer tutorials please use this facility.
3. Check that R studio will knit into an HTML document.
4. You can contribute the .rmd file via GitHub (https://github.com/jonesor/learn-basta), **to the master branch ONLY**. To do that you must be a collaborator on the repository (contact me with your username), or you can fork the repository and send a pull request (https://help.github.com/articles/using-pull-requests/).
5. Alternatively you can email me the *rmd file and associated files (e.g. any figures/diagrams.)
6. I will then compile the site on my computer and push it to the gh-pages branch (which is where the website lives). You could also do this, but you will need to have Jekyll and pandoc installed.
-->

```{r, echo=FALSE, results='hide', message=FALSE}
setwd("~/learn-basta")
library(BaSTA)
```

## Introduction {#intro}

In the previous section you learned how to format the data ready for analysis using `basta`. In this section we'll run through how to fit a simple Gompertz model - the default for BaSTA. This model will not have any covariates - we are simply going to estimate the coefficients of the fitted Gompertz model.

## Fitting a Gompertz model {#gompertz}

For this section, let's load the full dataset without errors... Read in the file 
``fixedData.csv``, which we will simply call ``dat``...

```{r}
dat <- read.csv("data/fixedData.csv", header = TRUE)
```

Now you have created a BaSTA-ready data set its time to fit a simple model. In fact, if you successfully finished the last session, you will already have fitted a BaSTA model.

To recap, this is how to do it...

```{r, cache=TRUE}
gsnc <- basta(dat, studyStart = 1970, studyEnd = 2000)
```

where, for simplicity we have named the BaSTA output object as `gsnc` which stands for _Gompertz, simple, no covariates_.

You will notice that there are only three arguments to this function: the BaSTA data object, and the start and end years of the study. There are of course other arguments to the `basta()` function (see `?basta`). The most important ones concern the model structure. The default model, which the above code fits, is a standard Gompertz model with a `simple` shape. We will discuss the other options in a later section. 

## Visualising the BaSTA outputs {#visualising-outputs}

With BaSTA you can extract the outputs from the BaSTA objects in different ways. You can simply type the name of the object and the built-in function `print()` will recognise the object and produce the following output

```{r}
gsnc
```

To get more information, simply use function `summary()` like this:

```{r}
summary(gsnc)
```

In addition, the BaSTA object is recognised by function `plot()`, allowing you to plot different aspects of the outputs. For instance, to visually determine if the MCMC parameter chains or "traces" have converged, simply type:


```{r, fig.height=7,fig.width=4}
plot(gsnc)
```

which produces the trace plots for the mortality parameters. To make sure that the recapture probability has converged, use argument `trace.name = "pi"` as

```{r, fig.height=4,fig.width=4}
plot(gsnc, trace.name = "pi")
```

To plot the mortality and survival trajectories, use argument `plot.trace = FALSE`, which produces the following plot:

```{r, fig.height=7,fig.width=4}
plot(gsnc, plot.trace = FALSE)
```

## Fancy plots {#fancy-plots}
Finally, you can produce what we call a "fancy plot" that combines the densities of the mortality parameters and the trajectories of mortality and survival into the same plot as

```{r, eval=FALSE, fig.height=7,fig.width=4}
plot(gsnc, fancy = TRUE)
```


